---
title: "Extraction/Fusion des données diatomiques des bases NAIADE et PANDORE"
author: "Léonard Heinry"
date: "16/08/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Packages loading 


```{r message = FALSE, warning = FALSE}
library(knitr)
require(tidyverse)
require(InraeThemes)
require(readxl)
require(fuzzyjoin)
library(readr)
library(leaflet)
library(sf)
require(tictoc)
library(lubridate)
library(purrr)
library(data.table)
library(progress)
library(rlist)
library(openxlsx)
library(purrr)
library(cooccur)
library(magrittr)
library(RODBC)
library(RMySQL)
library(RPostgreSQL)
```

## Diatoms data 

The first step of this work consist in getting all diatoms sampling that have 
been made since 2007. To do so, we will extract data from NAIADES and PANDORE
databases.


### NAIADES data 

```{r message = FALSE, warning = FALSE}
Diatom <- as.tibble(fread("data/Donnes_biologiques/Naiades_Export_France_Entiere_HB_2022/fauneflore.csv")) %>% # Data loading

  # We rename columns in order to standardize datasets 
  select("CODE_STATION" = CdStationMesureEauxSurface,
         "CODE_OPERATION" = RefOperationPrelBio,
         "Nom_groupe_taxo" = LbSupport,
         "DATE" = DateDebutOperationPrelBio,
         "SANDRE" = CdAppelTaxon,
         "Nom_latin_taxon" = NomLatinAppelTaxon,
         "RESULTAT" = RsTaxRep,
         "Code_groupe_taxo" = CdSupport) %>%

  filter(Code_groupe_taxo == 10) %>% # we only keep Diatom data 
  arrange(DATE,
          CODE_STATION, CODE_OPERATION,
          Nom_latin_taxon, RESULTAT); kable(head(Diatom))
```

PANDORE data are a little more difficult to get as they are stored on a web data
base that require SQL langage. In order to perform SQL programmation in R, we 
need to create an online gateway.

### Pandore data

```{r message = FALSE, warning = FALSE}

# Creation of a weblink to extract SQL data from R
channel=odbcConnect("pandore")

# Date format modification
Diatom2 <- Diatom %>% mutate(DATE = as.Date(DATE)) %>% select(-Code_groupe_taxo, -Nom_latin_taxon, -Nom_groupe_taxo) %>% 
  
# Extraction of PANDORE dataset and joining NAIADES dataset
bind_rows(as.tibble(sqlQuery(channel, "SELECT cd_opecont, cd_taxon, cd_OMNIDIA, comptage, cd_site, date_opecont, x, y, commune FROM compil
         join pandore.site using (cd_site)
         join pandore.listes_diatomee using (cd_opecont) WHERE date_opecont BETWEEN '2007-01-01' AND '2021-12-31'") %>% # SQL langage to extract data from Pandore
           select(CODE_OPERATION = "cd_opecont", CODE_STATION = "cd_site",
                               DATE = "date_opecont", SANDRE = "cd_taxon", CODE_TAXON = "cd_OMNIDIA", RESULTAT
                  = "comptage", x, y, commune)) %>% 
           mutate(CODE_OPERATION = as.character(CODE_OPERATION))) %>% 
  arrange(DATE) %>%
  # Equal rows between both dataset are removed as they represent the same sample 
  distinct(.keep_all = T); kable(head(Diatom2))
```

Now that the data is attached, we will add a way to identify the taxa in the database 
using the OMNIDIA code that is unique to each diatom species. 
These codes have been retrieved upstream in an excel file that we will
join to the Diatom2 file

### Taxa identification codes
```{r message = FALSE, warning = FALSE}
# Codes are coming from different sources to make sure we don't miss one
code <- bind_rows(as.tibble(read.csv2("data/Code_4L_taxons/Codes 4 lettres taxons.csv", stringsAsFactors = FALSE) %>% 
                              mutate(SANDRE = as.numeric(SANDRE))) %>% 
                    select(AFNOR = "cd_apa", "SANDRE"),as.tibble(read.csv2("data/Code_4L_taxons/Base_Omnidia.csv", stringsAsFactors = FALSE) %>% 
                                                                   select(AFNOR = CODE, "SANDRE") %>% 
                                                                   filter(SANDRE!=0)  %>% 
                                                                   mutate(SANDRE = as.numeric(SANDRE))), as.tibble(read.csv2("data/Code_4L_taxons/Complement_code4L.csv", stringsAsFactors = FALSE)) %>% 
                    select(AFNOR = "TAXON", "SANDRE") %>% 
                    mutate(SANDRE = as.numeric(SANDRE)) %>% drop_na()) %>% 
  distinct(SANDRE, .keep_all = T) %>%  
  bind_rows(data.frame(AFNOR = c("ERMS","FGNO","DIAS","CMFO","LSAP"),
                       SANDRE = c(4767, 38468, 20336, 45516, 66512))); kable(head(code))
```

What we need to be careful of is that in new taxonomy presents codes that 
are no longer the same as before. Depending on the sampling date, we might then 
have different codes for the same taxon. To avoid that, we will use a 
transcodification table to update all codes in order to get their latest version.

### Transcodification 
```{r message = FALSE, warning = FALSE}
# Joining codes
Transcoded_Flore <- Diatom2 %>% left_join(code, by = "SANDRE") %>% 
  mutate(AFNOR = if_else(is.na(AFNOR), CODE_TAXON, AFNOR)) %>% 
  select(DATE, CODE_STATION, AFNOR, SANDRE, RESULTAT, x, y, commune) %>% 
  filter(SANDRE != 0) %>% 
  # Joining updated codes 
  left_join(as.tibble(read.csv2("data/TableTranscodification.csv", stringsAsFactors = FALSE)) %>% 
              select(AFNOR = "ABRE", True_name = "CodeValid"), by = "AFNOR") %>%
  # If the code is the same as the updated then we keep it
  # if not we change it
  mutate(AFNOR = if_else(is.na(True_name) == T, AFNOR, True_name)) %>% 
  select(-True_name) %>% filter(!is.na(AFNOR)) %>% rename(code_taxon = "AFNOR"); kable(head(Transcoded_Flore))
```

At last, PANDORE data have stations coordinates directly included in samples but
NAIADES data dont. To standardize the file and make it easier to analyse then 
we are going to get missing coordinates from a station file we downloaded earlier

### Missing coordinates recovery

```{r message = FALSE, warning = FALSE}
Diatom3 <- Transcoded_Flore %>% mutate(CODE_STATION = str_remove(CODE_STATION, "^0+")) %>% 
  left_join(as.tibble(read.csv2("data/Donnes_traitees/Station.csv", stringsAsFactors = FALSE)) %>% 
              select(CODE_STATION = CdStationMesureEauxSurface, x = CoordXStationMesureEauxSurface, 
                     y = CoordYStationMesureEauxSurface, Commune = LbCommune) %>% 
              mutate(x = as.numeric(x), y = as.numeric(y)), by = "CODE_STATION") %>% 
  mutate(x = if_else(is.na(x.x == T), x.y, x.x),
         y = if_else(is.na(y.x == T), y.y, y.x),
         Commune = if_else(is.na(Commune == T), commune, Commune),
         Commune = if_else(is.na(Commune==T), "Inconnue", Commune)) %>% 
  # Sample without coordinates are unusable, we remove them (drop_na())
  select(-x.x, -y.x, -x.y, -y.y, -commune) %>% drop_na(); kable(head(Diatom3))
```


The diatom file is ready, now we need to get corresponding chemical samples. 
On a diatom station, one sample was made per year whereas one sample per month
was made for chemical parameters. Thus, for one diatom list of a station at
a given date, we need to extract madian of chemical parameters of interest 60 
days before sampling and 15 days after, so that environnemental parameters are 
relevant of tha conditions the diatoms lived before being sampled.

## Chemical samples 

### PANDORE chemical database

```{r message = FALSE, warning = FALSE}
# Link creation to use SQL programmation in R
channel=odbcConnect("pandore")
```


```{r message = FALSE, warning = FALSE}
# We only extract parameters of interest (1304,1311,1302,1314,1335,1433,1340)
Chimie_pandore <- as.tibble(sqlQuery(channel, "SELECT cd_opecont, cd_site, date_opecont, cd_obsPhyChi, cd_support, resultat, cd_param, cd_unite, nom_param FROM sandre_parametre
    join pandore.listes_phychi using(cd_param)
    join pandore.opecont_phychi using(cd_obsPhyChi)
    join pandore.compil using(cd_opecont)
    WHERE cd_param = 1302 OR cd_param = 1304 OR cd_param = 1311 OR 
                        cd_param = 1314 OR cd_param = 1335 OR cd_param = 1433 OR cd_param = 1340 AND cd_support = 3") %>% 
                      rename(CODE_STATION = cd_site,
                             Code_Unite_mesure = cd_unite,
                             Nom_parametre = nom_param,
                             Code_parametre = cd_param,
                             DATE = date_opecont,
                             Concentration = resultat,
                             Code_Prelevement = cd_obsPhyChi,
                             CdSupport = cd_support) %>% 
                      select(-cd_opecont) %>% 
                      arrange(DATE) %>% 
                      subset(CODE_STATION %in% c(unique(Diatom3$CODE_STATION))) %>% 
                      mutate(Code_Unite_mesure = as.character(Code_Unite_mesure),
                             Code_Prelevement = as.character(Code_Prelevement)) %>% 
                      select(CODE_STATION, Code_Prelevement, DATE,       
                             Code_parametre, Nom_parametre,
                             Concentration, Code_Unite_mesure))

head(Chimie_pandore)
```

### NAIADES chemical database

Chemical samples from NAIADES database were dowloaded earlier from there website 
just as we did for diatom data. NAIADES files are stored on computer as: 

"data/Donnes_chimiques/Naiades_Export_France_Entiere_PC_2022/analyses_2007.csv"
"data/Donnes_chimiques/Naiades_Export_France_Entiere_PC_2022/analyses_2008.csv"
etc...
                                    
## Diatoms and chemical data fusion                                    

To fuse data, we prepared a function that we can source directly from computer.
This function allow to match chemical samples 60 days before and 15 days after 
diatom samples according to station and year.

```{r message = FALSE, warning = FALSE}
source("R/Fonction_fusion_chimie_biologie.R")
```
 
In order to avoid binding PANDORE and NAIADES data and binding corresponding
chemical data to diatom samples for each year, we have decided to build another
function that directly implement Fonction_fusion_chimie_biologie.R 
after binding PANDORE and NAIADES chemical data. 

The function will need a date and will extract the required diatom and chemical
data to apply the fusion function and write an excel file so that we can save 
the results.


```{r message = FALSE, warning = FALSE}
source("R/Fonction_pick_date.R")
```

As amount of data is huge, we will not run the function but we will juste show 
the results 

```{r message = FALSE, warning = FALSE}

# What we are supposed to run

# Tab_2007 = Pick_date(2007)
# Tab_2008 = Pick_date(2008)
# Tab_2009 = Pick_date(2009)
# Tab_2010 = Pick_date(2010)
# Tab_2011 = Pick_date(2011)
# Tab_2012 = Pick_date(2012)
# Tab_2013 = Pick_date(2013)
# Tab_2014 = Pick_date(2014)
# Tab_2015 = Pick_date(2015)
# Tab_2016 = Pick_date(2016)
# Tab_2017 = Pick_date(2017)
# Tab_2018 = Pick_date(2018)
# Tab_2019 = Pick_date(2019)
# Tab_2020 = Pick_date(2020)
# Tab_2021 = Pick_date(2021)

# Example of Results 

kable(head(read.xlsx("data/Donnes_traitees/2008_traitee.xlsx") %>% 
  mutate(DATE = convertToDate(DATE)) %>% select(-Code_Prelevement)))

```

We get one dataset per year that we can use to map repartition of taxa of interest
However, statistical studies we want to conduct require full dataset. We thus
need to merge all the years.   

## Data gathering

```{r message = FALSE, warning = FALSE}
# We load all datasets from 2007 to 2021 in a list

x <- as.tibble(list(file = c("data/Donnes_traitees/2007_traitee.xlsx", "data/Donnes_traitees/2007_traitee.xlsx",
                             "data/Donnes_traitees/2008_traitee.xlsx", "data/Donnes_traitees/2008_traitee.xlsx",
                             "data/Donnes_traitees/2009_traitee.xlsx", "data/Donnes_traitees/2009_traitee.xlsx",
                             "data/Donnes_traitees/2010_traitee.xlsx", "data/Donnes_traitees/2010_traitee.xlsx",
                             "data/Donnes_traitees/2011_traitee.xlsx", "data/Donnes_traitees/2011_traitee.xlsx",
                             "data/Donnes_traitees/2012_traitee.xlsx", "data/Donnes_traitees/2012_traitee.xlsx",
                             "data/Donnes_traitees/2013_traitee.xlsx", "data/Donnes_traitees/2013_traitee.xlsx",
                             "data/Donnes_traitees/2014_traitee.xlsx", "data/Donnes_traitees/2014_traitee.xlsx",
                             "data/Donnes_traitees/2015_traitee.xlsx", "data/Donnes_traitees/2015_traitee.xlsx",
                             "data/Donnes_traitees/2016_traitee.xlsx", "data/Donnes_traitees/2016_traitee.xlsx",
                             "data/Donnes_traitees/2017_traitee.xlsx", "data/Donnes_traitees/2017_traitee.xlsx",
                             "data/Donnes_traitees/2018_traitee.xlsx", "data/Donnes_traitees/2018_traitee.xlsx",
                             "data/Donnes_traitees/2019_traitee.xlsx", "data/Donnes_traitees/2019_traitee.xlsx",
                             "data/Donnes_traitees/2020_traitee.xlsx", "data/Donnes_traitees/2020_traitee.xlsx",
                             "data/Donnes_traitees/2021_traitee.xlsx", "data/Donnes_traitees/2021_traitee.xlsx"),
                    sheet = c("chimie", "flore", "chimie", "flore",
                              "chimie", "flore", "chimie", "flore",
                              "chimie", "flore", "chimie", "flore",
                              "chimie", "flore", "chimie", "flore",
                              "chimie", "flore", "chimie", "flore",
                              "chimie", "flore", "chimie", "flore",
                              "chimie", "flore", "chimie", "flore",
                              "chimie", "flore")))


# We read flora and chemical sheets independently
Files <- map2(x$file, x$sheet, ~ read_xlsx(path = .x, sheet = .y))

# We seperatly store chemical and flora sheets, chemical = odd, flore = even
chimie_sheet = seq(1,30,2)
flore_sheet = seq(2,30,2)
```


```{r message = FALSE, warning = FALSE}
# Merge all chemical sheets together
Chimie <- bind_rows(Files[chimie_sheet]) %>%
  select(CODE_STATION,DATE,Code_Prelevement,
         Code_parametre,Nom_parametre,Code_Unite_mesure,Mediane)

kable(head(Chimie))
```


```{r message = FALSE, warning = FALSE}
# Merge all flora sheets together
Flore <- bind_rows(Files[flore_sheet]) %>%
  select(CODE_STATION, DATE, code_taxon, SANDRE, RESULTAT, Commune, x, y) %>%
  group_by(CODE_STATION, DATE) %>%
  mutate(Tot_indiv = sum(RESULTAT), RESULTAT = (RESULTAT*1000)/Tot_indiv) %>%
  ungroup() %>% mutate_if(is.numeric, round, 0)

kable(head(Flore))
```





